Data Specialization for a Staged Lambda Calculus

### 2015-02-25 paper/intro.tex
minor edits

\newcommand{\langname}{LAMDA12}
\section{Introduction}

@@

Partial evaluation performs precomputation by distinguishing between a
{\em static} and {\em dynamic} stage and performs all static
computations first, followed be the dynamic stage.  For example,
suppose that we wish to compute $F(x, y_0 \ldots y_{m-1})$ for a
statically known $x = 0$.  We can derive a specialization of $f$ at
$0$, written $f_0$, such that $f(x,y) = f_{0}(y)$ and specialize $F$
for $x=0$ as
\begin{lstlisting}
fun @$F_{0}(y_0, y_1 \ldots y_{m-1}$@ = 
  @$(f_{0}(y_0);  f_{0}(y_1)  \ldots f_{0}(ym)$@.
\end{lstlisting}
%
Such a partial evaluation typically proceed by performing a
binding-time analysis to determine the static and dynamic parts of $f$
and $F$ and specializes them based on the results of the binding-time
analysis.

@@

The third and final approach, called {\em pass separation} or {\em
  data specialization}, assumes much less information: instead of
requiring the static data, it requires just the knowledge of which
computations are static and dynamic and uses this knowledge to
generate two stage computations. 
%
Applied to our example, pass separation would {\em split} the function
$f$ into $f_s$ (static) and $f_d$ (dynamic) such that $f(x,y) =
f_d(f_s(x),y)$.  We can then rewrite our example as
%
\begin{lstlisting}
fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
let @$fsx = f_s(x)$@
in @$(f_d(fsx, y_0);  f_d(fsx, y_1)  \ldots f_d(fsx, y_{m-1})$@.
\end{lstlisting}
We refer to such a program as {\em stratified} program.

The key difference with partial evaluated program and the stratified
program is that the latter has lost to generality: it work for any
arguments, whereas the partially evaluated program is specialized for
a particular value of $x$ ($x = 0$ in the example).  Thus, the pass
separation can be applied without leading to any loss of generality.
Going back to the terminology of Jorring and Scherlis, pass separation
is a frequency reduction, whereas as partial evaluation and
meta-programming are precomputation techniques.  Specifically, if
desired partial evaluation can be applied further to a stratified
program as an orthogonal improvement.

But what is the benefit or pass separation? Let's assume for
concreteness that $f(x,y)$ takes time in the size of the first
argument.  The function $F(x, y_0 \ldots y_{m-1}$ then takes time $O(n
\cdot m)$ where $n$ is the size of $x$. Assume further that in the
stratified code the static function $f_s(x)$ takes $O(n)$ time and the
$f_d(fsx, \cdot)$ function takes $O(\log{n})$ time.  With these run
times, the stratified code has complexity $O(n + m\log{n})$ instead of
$O(n \cdot m)$, a significant asymptotic improvement.  (In this paper,
we will achieve such a reduction in complexity automatically.)

In summary pass separation causes no loss of generalization and can
improve efficiency, so why is it not applied liberally?

Indeed, in fact it is: in computer graphics, program are typically
manually stratified for frequency reduction.  In fact, stratification
into a pre-determined set ``modules'' is required by the execution
model of modern graphics architectures.  Such manual stratification is
a massive challenge: it breaks composition because the invariants are
scattered over the code written over multiple strata and because
modification to any one stratum requires modifying all others to
match.  Ideally, we wish to write just one program and stratify it
into multiple strata automatically.

Such automatic stratification turns out to be challenging: the first
successful approach that we know of is by Knoblock and
Ruf\,\cite{knoblock96}, which like our work was inspired by computer
graphics applications.  While practically effective, Knoblock and Ruf
consider only a very simple language and are not able to stratify
recursive programs.  Recent
efforts\,\cite{Proudfoot:2001,Foley:2011,He:2014} .... 
but still cannot stratify recursive programs.
%
In fact back in 80,s Jorring and Sherlis was able to stratify
reasonably complex programs but only manually and note that
``... partial evaluation technique is readily mechanized, while
staging transformation approach will elude full automation for some
time.''

In this paper we make significant progress on the problem of automatic
stratification of higher-order programs.  As our starting point, we
take the modal language \langname with explicit staging in the style
of Davies~\cite{Davies}, where the ``circle'' modality denotes
computation in a later stage, and present a staged operational
semantics similar for \langname (\secref{language}).  We then present
a splitting algorithm that stratifies a \langname program into a two
mono-staged programs, one for the first stage and for the second stage
(\secref{splitting}).  Since each program is at a mono-stage, they are
expressed in conventional function languages.

The crux of the splitting algorithm is splitting recursive mixed stage
functions that involve both a first stage and second stage
computations.  When split, such a function turns into a first-stage
function that yields a {\em boundary value} that encodes the
``first-stage'' part of the computation.  If the function recurs on a
first-stage argument, then the boundary value is typically a recursive
data structure.  In other words, the splitting algorithm maps
computational recursion into a recursive data type. If the boundary
structure is generated as a result of a conditional, it is typically
encoded as a (recursive) sum type, such as a tree, where the kind of
the node indicates the recursion status of the first-stage evaluation.
The ``second stage'' part of the computation takes the boundary value
as an argument as well as the second stage argument and uses them to
complete the computation, often by traversing the now precomputed
boundary in light of the now available second-stage argument.

Since it deals with unrestricted recursion and a rich language
consisting of first-class functions, recursion, and sum and recursive
types, the splitting algorithm is naturally complex. With some care,
the algorithm nevertheless can be specified reasonably succinctly.  In
fact, we have implemented the algorithm (\secref{implementation}) and
applied to several examples (\secref{examples}).  As an interesting
example, we show that a function 
\begin{lstlisting}
fun @$F(x: int list, y_0:int, y_1:int \ldots y_{m-1}:int$@ = 
  @$f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@,
\end{lstlisting}
where $f(x,y)$ selects the element of the list $x$ with rank $y$, when
splits yields an asymptotically more efficient function that runs in
expected $O(n\log{n} + m\log{n})$ time instead of expected $O(n \cdot
m)$ time, a near linear time improvement.  Interestingly the code
output by our splitting algorithm auto-generates the ``quicksort''
algorithm, which it uses to build a binary search tree as a boundary
value in the first stage, and in the second stage, generated a
``binary search'' algorithm that takes the tree and performs a binary
search on the tree as guided by the given rank.  To the best of our
knowledge, no prior approaches to this problem can perform such
complex transformations on code that can yield asymptotic
improvements in run time.

>>>>

\newcommand{\langname}{LAMDA12\xspace}
\section{Introduction}

@@

Partial evaluation performs precomputation by distinguishing between a
{\em static} and {\em dynamic} stage and performs all static
computations first, followed be the dynamic stage.  For example,
suppose that we wish to compute $F(x, y_0 \ldots y_{m-1})$ for a
statically known $x = [~]$.  We can derive a specialization of $f$ at
$[~]$, written $f_{[~]}$, such that $f(x,y) = f_{[~]}(y)$ and specialize $F$
for $x=[~]$ as
\begin{lstlisting}
fun @$F_{[~]}(y_0, y_1 \ldots y_{m-1})$@ = 
  @$(f_{[~]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
\end{lstlisting}
%
Such a partial evaluation typically proceed by performing a
binding-time analysis to determine the static and dynamic parts of $f$
and $F$ and specializes them based on the results of the binding-time
analysis.

@@

The third and final approach, called {\em pass separation} or {\em
  data specialization}, assumes much less information: instead of
requiring the static data, it requires just the knowledge of which
computations are static and dynamic and uses this knowledge to
generate two stage computations. 
%
Applied to our example, pass separation would {\em split} the function
$f$ into $f_s$ (static) and $f_d$ (dynamic) such that $f(x,y) =
f_d(f_s(x),y)$.  We can then rewrite our example as
%
\begin{lstlisting}
fun @$F(x, y_0, y_1 \ldots y_{m-1})$@ = 
let @$fx = f_s(x)$@
in @$(f_d(fx, y_0);  f_d(fx, y_1)  \ldots f_d(fx, y_{m-1}))$@.
\end{lstlisting}
We refer to such a program as {\em stratified} program.

The key difference with partial evaluated program and the stratified
program is that the latter has lost no generality: it works with all
arguments, whereas the partially evaluated program is specialized for
a particular value of $x$ ($x = [~]$ in the example).  Thus, the pass
separation can be applied without leading to any loss of generality.
In fact, this is quite natural: going back to the terminology of
Jorring and Scherlis, pass separation is a frequency reduction,
whereas as partial evaluation and meta-programming are precomputation
techniques. Consequently, if desired, partial evaluation can be
orthogonally applied to a stratified program.

% Let's assume for concreteness that $f(x,y)$ takes time in the size of
% the first argument.  The function $F(x, y_0 \ldots y_{m-1}$ then takes
% time $O(n \cdot m)$ where $n$ is the size of $x$. Assume further that
% in the stratified code the static function $f_s(x)$ takes $O(n)$ time
% and the $f_d(fsx, \cdot)$ function takes $O(\log{n})$ time.  With
% these run times, the stratified code has complexity $O(n + m\log{n})$
% instead of $O(n \cdot m)$, a significant asymptotic improvement.  (In
% this paper, we will achieve such a reduction in complexity
% automatically.)

But what is the benefit or pass separation? The big benefit is
efficiency. By stratifying the program, pass separation factors out
frequently used the parts of computation, computes them first, and
then reuse them later instead of computing them again.  Since it can
be applied statically with no loss of generality, pass separation can
significantly improve performance.  In fact, in computer graphics,
program are typically manually stratified both for improved
efficiency. Such stratification is in fact required by the execution
model of some modern graphics architectures.

While highly desirable, manual stratification is difficult and highly
nonmodular: program invariants are scattered over the code written in
multiple strata and because modification to any one stratum requires
modifying all others to match.  Ideally, we wish to write just one
program and stratify it into multiple strata automatically.  But such
automatic stratification turns out to be challenging: the first
successful approach that we know of is by Knoblock and
Ruf\,\cite{knoblock96}, which like our work was inspired by computer
graphics applications.  While practically effective, Knoblock and Ruf
consider only a very simple language and are not able to stratify
recursive programs.  Recent
efforts\,\cite{Proudfoot:2001,Foley:2011,He:2014} ....  but still
cannot stratify recursive programs.
%
In fact back in 80,s Jorring and Sherlis was able to stratify
reasonably complex programs but only manually and note that
``... partial evaluation technique is readily mechanized, while
staging transformation approach will elude full automation for some
time.''

In this paper we make significant progress on the problem of automatic
stratification of higher-order functional programs.  As our starting
point, we take the modal language \langname with explicit staging in
the style of Davies~\cite{Davies}, where the ``circle'' modality
denotes computation in a later stage, and present a staged operational
semantics similar for \langname (\secref{language}).  We then present
a splitting algorithm that stratifies a \langname program into a two
mono-staged programs, one for the first stage and for the second stage
(\secref{splitting}).  Since each program is at a mono-stage, they are
expressed in conventional function languages.

The crux of the splitting algorithm is splitting recursive mixed stage
functions that involve both a first stage and second stage
computations.  When split, such a function turns into a stratum-1
function that yields a {\em boundary value} that encodes the
``first-stage'' part of the computation.  If the function recurs on a
first-stage argument, then the boundary value is typically a recursive
data structure.  In other words, the splitting algorithm maps
computational recursion into a recursive data type. If the boundary
structure is generated as a result of a conditional, it is typically
encoded as a (recursive) sum type, such as a tree, where the kind of
the node indicates the recursion status of the first-stage evaluation.
The stratum-2 part of the computation, which corresponds to the
``second stage'' takes the boundary value as an argument as well as
the second-stage argument and uses them to complete the computation,
often by traversing the now precomputed boundary in light of the now
available second-stage argument.

Since it deals with unrestricted recursion and a rich language
consisting of first-class functions, recursion, and sum and recursive
types, the splitting algorithm is naturally complex. With some care,
the algorithm nevertheless can be specified reasonably succinctly.  In
fact, we have implemented the algorithm (\secref{implementation}) and
applied to several examples (\secref{examples}).  As an interesting
example, we consider a function
\begin{lstlisting}
fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
  @$f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@,
\end{lstlisting}
where $f(x,y)$ selects the element of the list $x$ with rank $y$ by
using the ``quickselect'' algorithm.  When applied to this function,
our split algorithm yields an asymptotically more efficient function
that runs in expected $O(n\log{n} + m\log{n})$ time instead of
expected $O(n \cdot m)$ time---a near linear time improvement.
Interestingly the code output by our splitting algorithm
auto-generates the ``quicksort'' algorithm, which it uses to build a
binary search tree as a boundary value in the first stage.  For the
second stage, the splitting algorithm generates a ``binary search''
that takes the tree and performs a binary search on the tree as guided
by the given rank.  To the best of our knowledge, no prior approaches
can perform such complex transformations on higher-order code, nor can
they yield such asymptotic improvements in run time.
